Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution

### 2020-07-07 manuscript.Rmd
quick correction in manuscript file

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number from `r raw_ON %>% distinct(publication) %>% nrow()` publications, listed in the supplementary files 'ovariole_number_bibliography.pdf' and 'ovariole_number_bibliography.bib'. We matched these records to additional taxonomic information using the software TaxReformer[@TaxReformer]. For all subsequent analyses, we excluded observations made in non-reproductive individuals from eusocial species (workers), as well as two observations that represented significant outliers and could not be validated using additional sources[@su2015testicular;@hernandez2017biology]. See Supplemental Methods Section \ref{sec:gathering} for details.

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body weight[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogensis[@buning1994insect]. See Supplemental Methods Section \ref{sec:combining} for details.

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified.  Analyses of insect family-level ovariole number, egg size, and body size were performed using the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See Supplemental Methods Section \ref{sec:trees} for details.

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See Supplemental Methods Section \ref{sec:PGLS} for details.

Ancestral state reconstruction of oogenesis mode was performed with the R package corHMM[@beaulieu2013identifying], version `r packageVersion('corHMM')`, and models of trait evolution were compared using the R package Ouwie[@beaulieu2012modeling], version `r packageVersion('OUwie')`. Ancestral state reconstruction and model comparison were repeated 100 times over a posterior distribution of trees and resampling data to account for variation across records for the same taxon. See Supplemental Methods Section \ref{sec:oogenesis-model}.

Other comparisons of model fit were performed using the R package geiger[@harmon2007geiger], version `r packageVersion('geiger')`, and validated using a parametric bootstrap with the R package arbutus[@pennell2015model], version `r packageVersion('arbutus')`. See Supplemental Methods Section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM[@rabosky2014automatic], version 2.5.0. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in\ref{fig:BAMM-analysis-burnin} resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See Supplemental Methods Section \ref{sec:BAMM} for details.

@@

The dataset of insect ovariole number is available in the supplementary file 'ovariole_number_data.tsv' and has been uploaded to Dryad (doi:10.5061/dryad.59zw3r253). The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a.

@@

Given that this prediction is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships [@revell2009size]. Across insects we did not observe a significant relationship between ovariole number and egg size when controlling for body size (p-value `r all_table[6,]$'pval_max'`; Fig. 2b, \ref{fig:PGLS-Gilbert-name}, \ref{fig:PGLS-egg-ON-family}, Table \ref{tab:PGLS-posterior})), though we do note the presence of a non-significant negative trend, primarily driven by a few extreme values (e.g. wasps in the family Trigonalidae, which have many ovarioles and lay minute eggs, Fig. \ref{fig:PGLS-egg-ON-family}).

>>>>

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number from `r raw_ON %>% distinct(publication) %>% nrow()` publications, listed in the supplementary file 'ovariole_number_bibliography.pdf'. We matched these records to additional taxonomic information using the software TaxReformer[@TaxReformer]. For all subsequent analyses, we excluded observations made in non-reproductive individuals from eusocial species (workers), as well as two observations that represented significant outliers and could not be validated using additional sources[@su2015testicular;@hernandez2017biology]. See supplementary methods section \ref{sec:gathering} for details.

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body weight[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogensis[@buning1994insect]. See supplementary methods section \ref{sec:combining} for details.

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified.  Analyses of insect family-level ovariole number, egg size, and body size were performed using the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See supplementary methods section \ref{sec:trees} for details.

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See supplementary methods section \ref{sec:PGLS} for details.

Ancestral state reconstruction of oogenesis mode was performed with the R package corHMM[@beaulieu2013identifying], version `r packageVersion('corHMM')`, and models of trait evolution were compared using the R package Ouwie[@beaulieu2012modeling], version `r packageVersion('OUwie')`. Ancestral state reconstruction and model comparison were repeated 100 times over a posterior distribution of trees and resampling data to account for variation across records for the same taxon. See supplementary methods section \ref{sec:oogenesis-model}.

Other comparisons of model fit were performed using the R package geiger[@harmon2007geiger], version `r packageVersion('geiger')`, and validated using a parametric bootstrap with the R package arbutus[@pennell2015model], version `r packageVersion('arbutus')`. See supplementary methods section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM[@rabosky2014automatic], version 2.5.0. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in\ref{fig:BAMM-analysis-burnin} resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

The dataset of insect ovariole number is available at Dryad (doi:10.5061/dryad.59zw3r253). The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a.

@@

Given that this prediction is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships [@revell2009size]. Across insects we did not observe a significant relationship between ovariole number and egg size when controlling for body size (p-value `r all_table[6,]$'pval_max'`; Fig. 2b, \ref{fig:PGLS-Gilbert-name}, \ref{fig:PGLS-egg-ON-family}, Table \ref{tab:PGLS-posterior}), though we do note the presence of a non-significant negative trend, primarily driven by a few extreme values (e.g. wasps in the family Trigonalidae, which have many ovarioles and lay minute eggs, Fig. \ref{fig:PGLS-egg-ON-family}).
